### 2013-11-19 inst/stuff/taxize_vignette.Rmd

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{An R Markdown Vignette made with knitr}
-->

@@

```{r resolvenames, eval=TRUE, cache=TRUE}
temp <- gnr_resolve(names = c("Helianthos annus", "Homo saapiens"))
temp[ , -c(1,4)]
```

@@

```{r synonyms, eval=TRUE, cache=TRUE, message=FALSE, warning=FALSE, comment=NA}
mynames <- c("Helianthus annuus ssp. jaegeri", "Helianthus annuus ssp. lenticularis", "Helianthus annuus ssp. texanus", "Helianthus annuus var. lenticularis", "Helianthus annuus var. macrocarpus", "Helianthus annuus var. texanus")
tsn <- get_tsn(mynames)
ldply(tsn, itis_acceptname)
```

@@

```{r taxname, eval=TRUE,cache=TRUE}
tax_name(query = "Helianthus annuus", get = "family", db = "itis")
tax_name(query = "Helianthus annuus", get = "family", db = "ncbi")
```

@@

```{r interactive_many, eval=TRUE, echo=TRUE, cache=TRUE}
splist <- c("annona cherimola", 'annona muricata', "quercus robur",
  	"shorea robusta", "pandanus patina", "oryza sativa", "durio zibethinus")
get_tsn(searchterm = splist, searchtype = "sciname")
```

In another example, note that no match at all returns an NA:

```{r interactive_na, eval=TRUE, echo=TRUE, cache=TRUE}
get_uid(sciname = c("Chironomus riparius", "aaa vva"))
```

#### Retrieve a phylogeny

Ecologists are increasingly taking a phylogenetic approach to ecology, applying phylogenies to topics such as the study of community structure, ecological networks, functional trait ecology. Yet, Many biologists are not adequately trained in reconstructing phylogenies. Fortunately, there are some sources for getting a phylogeny without having to know how to build one; one of these is for angiosperms, called Phylomatic \cite{webb2005}. We have created a workflow in taxize that accepts a species list, and taxize works behind the scenes to get higher taxonomic names, which are required by Phylomatic to get a phylogeny. Here is a short example, producing the tree in figure (Fig. 1).

```{r phylomaticphylogeny_disp, eval=FALSE}
taxa <- c("Poa annua", "Abies procera", "Helianthus annuus")
tree <- phylomatic_tree(taxa = taxa)
tree$tip.label <- capwords(tree$tip.label)
plot(tree, cex=1)
```

Input the taxonomic names

```{r phylomaticphylogeny1, eval=TRUE, dev=c ('pdf','tiff'), dpi=300, out.width='4in', out.height='4in', cache=TRUE, include=FALSE}
taxa <- c("Poa annua", "Abies procera", "Helianthus annuus")
```

Fetch the tree - the formatting of names and higher taxonmy is done within the function

```{r phylomaticphylogeny2, eval=TRUE, dev=c ('pdf','tiff'), dpi=300, out.width='4in', out.height='4in', cache=TRUE, include=FALSE}
tree <- phylomatic_tree(taxa = taxa)
```

Captilize the species names

```{r phylomaticphylogeny3, eval=TRUE, dev=c ('pdf','tiff'), dpi=300, out.width='4in', out.height='4in', cache=TRUE, include=FALSE}
tree$tip.label <- capwords(tree$tip.label)
```

Plot the tree

```{r phylomaticphylogeny4, eval=TRUE, dev=c ('pdf','tiff'), dpi=300, out.width='4in', out.height='4in', cache=TRUE, include=FALSE}
plot(tree, cex=1)
```

Behind the scenes the function *phylomatic_tree} retrieves a Taxonomic Serial Number (TSN) from ITIS for each species name, then a string is created for each species like this `poaceae/oryza/oryza_sativa` (with format `family/genus/genus_epithet`). These strings are submitted to the Phylomatic API, and if no errors occur, a phylogeny in newick format is returned. The `phylomatic_tree()` function also cleans up the newick string and converts it to an **ape** `phylo` object. The output from `phylomatic_tree()` is a `phylo` object, which can be used for plotting and phylogenetic analyses. Be aware that Phylomatic has certain limitations - refer to the paper describing [Phylomatic][phylomatic] and the website [http://phylodiversity.net/phylomatic/][http://phylodiversity.net/phylomatic/].

There are currently no resources for getting a phylogeny of animals simply from species names. However, a few projects are working on this problem, including the [Open Tree of Life][opentree]. We will incorporate these resources when the appropriate APIs are available.

@@

#### IUCN Status

There are a number of things we can do once we have the correct taxonomic names. One thing we can do is ask about the conservation status of a species ([IUCN Red List of Threatened Species][iucn]). We have provided a set of functions, `iucn_summary` and `iucn_status`, to search for species names, and extract the status information, respectively. Here, we search for the Panther and *Lynx*.

```{r iucn, eval=TRUE, cache=TRUE}
ia <- iucn_summary(c("Panthera uncia", "Lynx lynx"))
iucn_status(ia)
```

It turns out that the panther has a status of endangered (EN) and the lynx has a status of least concern (LC).

#### Search for available genes in GenBank

Another use case available in taxize deals with genetic sequences. taxize has three functions to interact with GenBank to search for available genes (`get_genes_avail`), download genes by GenBank ID (`get_genes`), and download genes via taxonomic name search, including retrieving a congeneric if the searched taxon does not exist in the database (`get_seqs`). In this example, we search for gene sequences for *Umbra limi*.

```{r genbank_one, eval=TRUE, cache=TRUE}
out <- get_genes_avail(taxon_name = "Umbra limi", seqrange = "1:2000", getrelated = FALSE)
```

Then we can ask if *RAG1* exists in any of the gene names.

```{r genbank_two, eval=TRUE, cache=TRUE}
out[grep("RAG1", out$genesavail, ignore.case = TRUE),]
```

It turns out that there are 430 different unique records found. However, this doesn't mean that there are 430 different genes found as the API does not provide metadata to classify genes. However, you can use regular expressions (e.g., `grep`) to search for the gene of interest.

@@

#### Aggregating data to a specific taxonomic rank

In biology, one can asks questions at varying taxonomic levels. One may perform analyses on different taxonomic levels. This use case is easily handled in taxize. A function called `tax_agg` will aggregate community data to a specific taxonomic level. In this example, we take data of 5 species and aggregate them to family level. Again we can specify if we want to use data from ITIS or NCBI.

```{r tax_agg, eval=TRUE, cache=TRUE, tidy=FALSE}
data(dune, package = 'vegan')
df <- dune[ , 1:5]
colnames(df) <- c("Bellis perennis", "Empetrum nigrum", "Juncus bufonius", 
                  "Juncus articulatus", "xxx")
head(df)
agg <- tax_agg(df, rank = 'family', db = 'ncbi')
agg
head(agg$x)
```

[eol]: 
[taxosaurus]: 
[ncbi]: 
[itis]: 
[phylomatic]: 
[opentree]: 
[wikispecies]: 
[col]: 
[iucn]:

>>>>

<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{An R Markdown Vignette made with knitr}
-->

@@

```{r resolvenames, eval=TRUE, cache=TRUE}
library(taxize)
temp <- gnr_resolve(names = c("Helianthos annus", "Homo saapiens"))
temp[ , -c(1,4)]
```

@@

```{r synonyms, eval=TRUE, cache=TRUE, message=FALSE, warning=FALSE, comment=NA}
mynames <- c("Helianthus annuus ssp. jaegeri", "Helianthus annuus ssp. lenticularis", "Helianthus annuus ssp. texanus")
tsn <- get_tsn(mynames)
library(plyr)
ldply(tsn, itis_acceptname)
```

@@

```{r taxname, eval=TRUE,cache=TRUE}
tax_name(query = "Helianthus annuus", get = "family", db = "ncbi")
```

@@

```{r interactive_many, eval=TRUE, echo=TRUE, cache=TRUE}
splist <- c("annona cherimola", 'annona muricata', "quercus robur")
get_tsn(searchterm = splist, searchtype = "sciname")
```

@@

@@

[eol]: http://eol.org/
[taxosaurus]: http://api.phylotastic.org/tnrs
[ncbi]: http://www.ncbi.nlm.nih.gov/
[itis]: http://www.itis.gov/
[phylomatic]: http://phylodiversity.net/phylomatic/
[opentree]: http://blog.opentreeoflife.org/
[wikispecies]: http://species.wikimedia.org/wiki/Main_Page
[col]: http://www.catalogueoflife.org/
[iucn]: http://www.iucnredlist.org/
